A smart street lighting system based on IPv6-enabled wireless sensor network

### 2014-10-30 2014_ELSEVIER_CEE_SI/paper_cee_special_issue.tex
Cleanup
1st submission

\documentclass[final,authoryear,3p,twocolumn]{elsarticle}

%% Use the option review to obtain double line spacing
%\documentclass[authoryear,preprint,review,12pt]{elsarticle}

@@

\usepackage{pgfplots}
\usetikzlibrary{arrows,automata,petri}
\usetikzlibrary{shapes,snakes,positioning}

@@

\author{Carlos Henrique Barriquello, Gustavo Weber Denardin}
%\address{Electronic Ballast Researching Group (GEDRE) - Intelligence in Lighting, Federal University of Santa Maria(UFSM), Santa Maria, RS, 97105-900, Brazil}

\begin{abstract}
%% Text of abstract
In this paper, we propose propose a smart street lighting system based on IPv6-enabled wireless sensor network. We discuss the main drawbacks of RPL routing protocol for P2P communication in large-scale 6LoWPAN networks. Then, we address such drawbacks by proposing a new routing algorithm, named GeoRank, which integrates RPL protocol with GOAFR position-based routing protocol. Simulations were performed on networks extracted from real street maps. The obtained results show that GeoRank finds shorter routes than RPL in high link density conditions and GOAFR in low link density conditions. Thus, GeoRank shows to be adaptive to variable link densities found in large-scale networks. Further, GeoRank avoids the use of bandwidth-consuming control messages required in RPL, while being more scalable in terms of memory usage than storing-mode RPL.
\end{abstract}

@@

%Highlights
%
%We propose a smart street lighting system based on a IPv6-enabled wireless sensor networks.
%
%Main drawbacks of RPL protocol for P2P communication are discussed.
%
%We propose a new routing algorithm, termed GeoRank, which integrates RPL protocol with GOAFR geographic routing protocol.
%
%Simulations were performed on networks extracted from real street maps.
%
%Results show that GeoRank performs better in large-scale networks with variable link densities than GOAFR or RPL protocols alone.
%
%We discuss the scalability of GeoRank in terms of memory and bandwidth consumption in comparison to RPL protocol.

@@

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
Wireless Sensor and Actuator Networks  \sep Smart Street Lighting \sep Routing \sep IPv6 \sep 6LoWPAN

@@

The street lighting system of a city plays an important role on its infrastructure. It is responsible of guaranteeing the society security and living during nighttime, traffic safety, and city appearance. Most commonly used lamp types in street lighting are the high pressure discharge lamps, e.g.: mercury vapor lamps, high-pressure sodium (HPS) lamps, and metal halide (MH) lamps \citep{HID_Lamps_2008,HID_Lamps_2009,HID_Lamps_2010a,HID_Lamps_2010b}. However, with the efficiency and color quality of light emitting diodes (LEDs) steadily improving, LED lamps have been recently introduced as an alternative to replace the conventional lighting systems \citep{LED_conf_2009,LED_conf_2010,LED_jou_2010a,LED_jou_2010b}.

The use of LEDs in street lighting systems started to be reported in the last few years, and it is being considered a very good solution to replace the HID-based lighting systems, because of the longer lifetime, higher luminous efficiency in mesopic visual conditions, and higher CRI \citep{CRI_2008,CRI_2010}. Moreover, LED lamps may also be dimmed to reduce its power consumption. However, LED lamps are usually less cheaper than HID lamps, and thus become an important asset in the system. Therefore, the presence of LED lamps in the modern street lighting systems brings a need for better management, in order to decrease the maintenance cost and power consumption (e.g.: by lamp dimming).

The addition of communication capability to the street lighting system goes towards such a goal. With a communication network, one can monitor the status information of each lamp (state, current, voltage and power measurements, etc.) and have better control of the system (e.g.: dimming). 
If the system is connected to a network, the lamp power can be monitored and the luminary can be quickly replaced when failure happens, because a decrease of active power consumption is observed. Not only a preventive maintenance of the system can be easily programmed, but also software updates can be done (which allows future enhancements and new functions addition). Then, a street lighting system with communication capability is justified by its advantages.

However, the street lighting systems features a large number of independent devices, with geographic distribution depending on the city streets. Then, adding communication capabilities to these devices requires a complex network topology, as well as interoperability, scalability, security, robustness, low cost, ease to use and maintenance. A few contributions in the literature suggest some technologies that can be used to control a street lighting system. Some use the power lines for data transmission (PLC) \citep{PLC_Street_Lighting_2003,PLC_Street_Lighting_2006}, while others use wireless communication, such as cellular networks \citep{Cellular_Street_Lighting_2009_1,Cellular_Street_Lighting_2010_1}, wireless sensor networks \citep{WSN_Street_Lighting_2010_1, WSN_Street_Lighting_2010_2} and both combined \citep{WSN_Street_lighting_2006, WSN_Street_lighting_2007}.

@@

Recently, IETF has been standardized 6LoWPAN protocol (RFC4944) as an adaptation layer to enable IEEE 802.15.4-based low power and lossy networks (LLNs), such as wireless sensor networks, to reuse existing IPv6 protocol for networking layer. The IP protocol is an open, mature and field proven technology, with free and well documented standards (RFCs), and thus it is very reasonable to reuse it for wireless sensor networking instead of creating new proprietary protocols. However, 6LoWPAN protocol does not define the routing protocol, which shall be defined apart. Thus, IETF has standardized RPL as the routing protocol for IP-based LLNs.

However, very recent empirical studies showed that RPL has some drawbacks, as higher amount of control messages than data messages, mainly for constructing routes from the sink to the nodes (Downward Routes in RPL vocabulary) \citep{Heurtefeux_RPL_2013}. By the other hand, in a street light system, nodes are static and a routing protocol could benefit of this condition to find routes with less control overhead than RPL. The routing algorithms that use position to find routes in a communication network are known as geographic routing algorithms and a large body of literature has been devoted to the research on those algorithms in last few decades \citep{Karp_GPSR_2000, Position_Based_Routing_Ad_Hoc_2001, On_Optimal_Geographic_Routing_2007, Denardin2011}.

%Paper outline:
%
%Introduction to the street lighting system and problems (0.5 pg) (OK)
%%-	Energy usage, lamps vs efficiency, HPS lamps, LED lamps, etc?
%%-	Required features of smart street lighting systems: energy/power measurement, lamp control (on/off/dimming), system diagnostics, failure report, etc?
%%-	The smart street lighting wireless node:  block diagram of smart street lighting wireless node and prototype (isso talvez poderia ser colocado como resultado)
%
%3) Network structure: base stations and router nodes (0.5 pg)
%3a) routing layer: why geographical routing? geographical routing  algorithms (1pg)
%3b) Network stack with 6LoWPAN, why use 6LoWPAN (0.5 pg)
%3c) Routing algorithms in 6LoWPAN: mesh under x route over. Existing protocols (RPL, DYMO, LOAD, etc..) (1.5 pg). Why not use these? 
%3d) proposal for geographic routing support in 6LoWPAN  (1 pg).  Compare to SPEED or to RPL? %(http://ieeexplore.ieee.org/stamp/stamp.jsp?tp=&arnumber=6058995)
%
%4) results and discussion of design choices, code/data footprint, processing time, data structures, algorithm implementation, etc? (0.5 pg)
%
%5) conclusions: opportunities, challenges and future research directions (0.5 pg)

@@

6LoWPAN is a protocol definition, standardized in RFC4944 and RFC6282, that works as an adaptation layer between the IPv6 network layer and the IEEE 802.15.4 MAC (Media Access Control) layer, as depicted in figure \ref{stack_6lowpan}. It is required because IEEE 802.15.4 standard, which defines MAC and PHY layers for low power and low data rate wireless networks, is able to carry packets with maximum size of 127 bytes, while IPv6 packets requires a MTU (minimum transmission unit) of 1280 bytes. Thus, the overall goal of 6LoWPAN is to allow IPv6 packets to be carried on top of IEEE 802.15.4-based wireless sensor networks, by defining fragmentation and compression mechanisms of the IPv6 packets and thus enabling IPv6-based wireless sensor networks \citep{Mulligan_6LoWPAN_2007}.

@@

\subsection{Routing in 6LoWPAN}

In order to address the lack of a routing protocol for 6LoWPAN networks, the IPv6 Routing Protocol for Low-Power and Lossy Networks (RPL) has been proposed in RFC6550 by IETF. Accordingly, RPL provides support for point-to-point, point-to-multipoint and multipoint-to-point traffic flows.

The RPL protocol is based on directed acyclic graph (DAG). DAGs are directed graphs formed by a collection of vertices and directed edges, where each edge connects one vertex to another, such that there is no way to start at some vertex v and follow a sequence of edges that eventually loops back to v again. All edges are contained in paths oriented toward and terminating at one or more root nodes. These root nodes are known as Low power and Lossy Network Border Routers (LBR), which are devices that connects the LLN to another routing domain such as a Local Area Network (LAN) or the internet, where a different routing protocol may be in operation. In addition to its functionality as a routing device, the LBR acts as a sink node to a WSN as well as a data collector or aggregator.

In RPL, a DAG grounded at a single LBR is called a Destination Oriented Directed Acyclic Graph (DODAG). For a specific DODAG each node is assigned to a rank such that rank increases as the node is far from the LBR. Note that the rank of the nodes along any path to the LBR must be monotonically decreasing, avoiding the routing loops, as shown in the figure \ref{RPL_DIO}. Therefore, forwarding a packet to the LBR roughly consist in choosing the neighbor node with the lowest rank.

\begin{figure}[h!tb]
\centering
\includegraphics[width=0.4\textwidth]{figures/RPL_DIO}
\caption{Example of a Destination Oriented Directed Acyclic Graph(DODAG).}
\label{RPL_DIO}
\end{figure}

Nodes running RPL exchange control messages to setup and maintain the DODAG, which in the 6LoWPAN implementation are ICMPv6 messages. The 1B code in the ICMPv6 header is used to differentiate between different sub-type of control messages. The sub-type used to build the DODAG is called DAG Information Object (DIO). All the nodes in the network regularly broadcast a DIO to its one-hop neighbors, indicating their activity and rank. The instant at which a node issues a DIO is governed by a Trickle timer, which is reset only when an inconsistency arises, i. e., very few DIOs are exchanged when the topology is stable. RPL provide optimized paths to one or more LBRs with very little information stored in the routing table. The information stored is limited to its one-hop neighbors and their parents in the DODAGs. A parent of a node within a DODAG is one of the immediate successors of the node on a path towards the DODAG root.

Despite of its advantages, the default RPL using DODAGs does not provide routes from the LBRs to other nodes in the network, which in RPL are known as down routes. Down refers to the direction from DODAG roots towards leaf nodes, following the reverse direction of DODAG edges. The availability of such routes is desirable in several scenarios, mainly in scenarios with actuators nodes and scenarios where selective queries to sensors nodes are required. Thus, a RPL router that requires a path from a LBR to itself should send a DAO (Destination Advertisement Object) message to the LBR through the DODAG edges. The nodes that receive a DAO message forward the message as soon as the timer DelayDAO expires, being this time determined by the implementation, according to [RFC6550]. Therefore, routes for any destination inside the network, other than the LBRs, are provided by these destinations through DAO messages.

@@

\begin{figure}[h!tb]
\centering
\includegraphics[width=0.45\textwidth]{figures/RPL_DAO}
\caption{P2P communication between routers $B$ and $C$ through a DODAG root using RPL.}
\label{RPL_DAO}
\end{figure}

@@

Even, there are problems related to the DAO operating modes. In non-storing mode, RPL requires the use of IPv6 source routing mechanism. Therefore, a Source Routing Header (SRH) for use strictly between RPL routers in the same RPL routing domain was proposed in RFC6554. However, RPL source routing implements a policy where each IPv6 hop between the source and destination of the source route is specified within the SRH. Such approach leads to a large overhead and increases the consumption of the already scarce bandwidth. Further, as the IEEE 802.15.4 standard has a limited maximum transmission unit (MTU), that may also imply in increased fragmentation. Therefore, successful delivery of an IP packet depends on the successful delivery of all possible fragments, which are not desirable once the LLNs are generally characterized by high packet loss rates.

In addition to the fragmentation issue, the maximum length of the routing header is limited to 255 octets [ref], which restricts the maximum path length to 127 hops if an address compression mechanism is applied [ref]. This limitation may exclude scenarios with long chain-like topologies, such as street lighting systems and traffic lights along a street.

In storing mode, each router has to be able to store all the routes downwards its position in the formed DODAGs. This implies that, for routers near to the root node, the required storage is only bounded by the number of paths to all other routers in the network. As RPL targets networks consisting of thousands of memory-constrained devices, the required storing capacity on these devices may not be compatible with the basic precepts of the WSANs.

@@

Geographic routing mainly relies on a really simple geographic greedy-forwarding strategy, where each router node must select a locally optimal neighbor with a positive progress towards the data packet destination \citep{survey_2007}. Such approach is largely known by its low overhead and scalability, which can easily be observed due to its low control message system, i.e., a router in a geographic routing strategy only needs to know its direct neighborhood.

@@

\section{GeoRank}

Mainly due to its greedy step, GOAFR algorithm is able to perform quite well in networks with high link density, discovering paths often optimal or near-optimal paths, in contrast to RPL, which is constrained to route only through the DODAGs and thus, being unable to find the optimal paths. By the other hand, in networks with low link density, mainly around the critical density (\citep{GOAFR_2008}, GOAFR has a performance loss and may usually find paths longer than RPL. Therefore, a hybrid approach that leverages the best of both algorithms according to the network link density is proposed. This approach is mainly suitable for large scale and heterogeneous networks, such as a network of wireless streetlight nodes, where the link density varies from low to high depending of the more or less presence of obstructions and buildings in different places of the network.  The operation of \textit{GeoRank} is illustrated in figure \ref{state_machine} and the algorithm is outlined below in Algoritm \ref{alg}.

\small
\begin{algorithm}
\begin{algorithmic}
\caption{GeoRank}
\label{alg}
\small
\STATE \textbf{Step 1} - Store in the packet the DODAG root with the smallest angle to the source-destination
line. 
\STATE \textbf{Step 2} - Start with greedy step of GOAFR until: (case I) finding a void node (i.e. a node where GOAFR would switch to face  routing (OAFR)) or (case II) delivering the packet. In case I, store the current position in the packet and go to step 3.
\STATE \textbf{Step 3} - If the destination node is closer to the DODAG root (stored in the packet) than the current node or the DODAG root is closer to the destination node than the current node, forward to the next hop in the DODAG towards the root (i.e. upward route). If neither condition is satisfied, go to step 4. Whenever the current node position is closer to destination than the position stored in the packet in step 2, go to step 1.
%\STATE \textbf{Step 3} - If the destination node is closer to the DODAG root (stored in the packet) than the current node(when source and destination nodes are in the same quadrant) or the DODAG root is closer to the destination node than current node (when source and destination nodes are in different quadrants), forward to the next hop in the DODAG towards the DODAG root (i.e. upward route). If neither condition is satisfied, go to step 4. Whenever the current node position is closer to destination than the position stored in the packet in step 2, go to step 1.
\STATE \textbf{Step 4} -  Execute face routing mode of GOAFR until delivering the packet or returning to step 1.
\small
\end{algorithmic}
\end{algorithm}
\normalsize
\begin{figure}
\centering
\begin{tikzpicture}
[scale=1, every path/.style={>=latex}, no/.style={draw,circle,align=center,text width = 1.3cm}]
  \node[no] (a) at (0,0)  {\small GOAFR greedy routing};
  \node[no] (b) at (3,4)  {\small DODAG upward routing};
  \node[no] (c) at (6,0) {\small GOAFR face routing };

@@

For every scenario, there were generated $1000$ random source-destination pairs, giving a coverage of up to $10$\% of the possible source-destination combinations. Also, every simulation was performed for $10$ different (random) positions of the DODAG root in every scenario, which gives a good coverage of the $5$ possible root placements (i.e. center and top/bottom-right/left corners). Then, for the each source-destination pair, the route was determined by each algorithm under test (RPL, GOAFR, \textit{GeoRank} and Dijkstra' algorithm for shortest path) and length of each route was measured in number of hops. The average length of the routes was computed and the results are shown in figure \ref{sim_results}.
%For RPL rank computation, it was used the Objective Function 0 as defined in RFC6552.
\begin{figure}
\centering
%\includegraphics[width=0.45\textwidth,natwidth=2024,natheight=402]{figures/sample_scenarios.eps}
\includegraphics[width=0.45\textwidth]{figures/sample_scenarios2.eps}
\caption{Sample of simulated scenarios with nodes populated.}
\label{sim_scenarios}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{figures/final_result.eps}
\caption{Simulation results of the average route length for the routing algorithms (RPL, GOAFR, \textit{GeoRank}, and Dijkstra' algorithm for shortest path).}
\label{sim_results}
\end{figure}

@@

\bibliographystyle{model2-names}
\bibliography{refs}

>>>>

%\documentclass[final,authoryear,3p,twocolumn]{elsarticle}

%% Use the option review to obtain double line spacing
\documentclass[authoryear,preprint,review,12pt]{elsarticle}

@@

\usepackage{pgfplots}
\usetikzlibrary{arrows,automata,petri}
\usetikzlibrary{shapes,decorations,positioning}

@@

\author[addr1]{Carlos Henrique Barriquello}
\author[addr2]{Gustavo Weber Denardin}
\author[addr1]{Alexandre Campos}
\address[addr1]{Electronic Ballast Researching Group (GEDRE) - Intelligence in Lighting, Federal University of Santa Maria(UFSM), Santa Maria, RS, 97105-900, Brazil}
\address[addr2]{Technological Federal University of Paran$\acute{a}$, Pato Branco, PR, 85503-390, Brazil}

\begin{abstract}
%% Text of abstract
In this paper, we propose a smart street lighting system based on IPv6-enabled wireless sensor network. We discuss the main drawbacks of RPL routing protocol for P2P communication in large-scale 6LoWPAN networks. Then, we address such drawbacks by proposing a new routing algorithm, named \textit{GeoRank}, which integrates RPL protocol with GOAFR position-based routing protocol. Simulations were performed on networks extracted from real street maps. The obtained results show that \textit{GeoRank} finds shorter routes than RPL in high link density conditions and GOAFR in low link density conditions. Thus, \textit{GeoRank} shows to be adaptive to variable link densities found in large-scale networks. Further, \textit{GeoRank} avoids the use of bandwidth-consuming control messages required in RPL, while being more scalable in terms of memory usage than storing-mode RPL.
\end{abstract}

@@

%Highlights
%
%We propose a smart street lighting system based on a IPv6-enabled wireless sensor networks.
%Main drawbacks of RPL protocol for P2P communication are discussed.
%We propose a new routing algorithm, termed GeoRank, which integrates RPL protocol with GOAFR geographic routing protocol.
%Simulations were performed on networks extracted from real street maps.
%Results show that GeoRank performs better in large-scale networks with variable link densities than GOAFR or RPL protocols alone.
%We discuss the scalability of GeoRank in terms of memory and bandwidth consumption in comparison to RPL protocol.

@@

%- We propose a smart street lighting system based on IPv6-enabled wireless sensor network.
%- Main drawbacks of RPL protocol for P2P communication are discussed.
%- GeoRank, a new routing algorithm is proposed and its scalability is compared to RPL.
%- Simulations were performed on networks extracted from real street maps.
%- GeoRank is adaptive to variable link densities found in large-scale networks.

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
Wireless Sensor and Actuator Networks  \sep Smart Street Lighting \sep Internet of Things \sep Smart Green Cities \sep Routing \sep IPv6 \sep 6LoWPAN

@@

The street lighting system of a city plays an important role on its infrastructure. It is responsible of guaranteeing the society security and living during nighttime, traffic safety, and city appearance. Most commonly used lamp types in street lighting are the high pressure discharge lamps, e.g.: mercury vapor lamps, high-pressure sodium (HPS) lamps, and metal halide (MH) lamps \citep{HID_Lamps_2010a,HID_Lamps_2010b}. However, with the efficiency and color quality of light emitting diodes (LEDs) steadily improving, LED lamps have been recently introduced as an alternative to replace the conventional lighting systems \citep{LED_jou_2010a,LED_jou_2010b}.

The use of LEDs in street lighting systems started to be reported in the last few years, and it is being considered a very good solution to replace the HID-based lighting systems, because of the longer lifetime, higher luminous efficiency in mesopic visual conditions, and higher CRI \citep{CRI_2008,CRI_2010}. Moreover, LED lamps may also be dimmed to reduce its power consumption. Therefore, the presence of LED lamps in the modern street lighting systems brings a need for better management, in order to decrease the maintenance cost and power consumption (e.g.: by lamp dimming). 
% However, LED lamps are usually less cheaper than HID lamps, and thus become an important asset in the system.

The addition of communication capability to the street lighting system goes towards such a goal. With a communication network, one can monitor the status information of each lamp (state, current, voltage and power measurements, etc.) and have better control of the system (e.g.: dimming). 
If the system is connected to a network, the lamp power can be monitored and the luminary can be quickly replaced when a failure occurs, because a decrease of active power consumption is observed. Not only a preventive maintenance of the system can be easily programmed, but also software updates can be done (which allows future enhancements and addition of new functions). Then, a street lighting system with communication capability is justified by its advantages.

However, a street lighting system features a large number of independent devices, with geographic distribution depending on the city streets. Then, adding communication capabilities to these devices requires a complex network topology, as well as interoperability, scalability, security, robustness, low cost, ease to use and maintenance. A few contributions in the literature suggest some technologies that can be used to control a street lighting system. Some use the power lines for data transmission (PLC) \citep{PLC_Street_Lighting_2003,PLC_Street_Lighting_2006}, while others use wireless communication, such as cellular networks \citep{Cellular_Street_Lighting_2009_1,Cellular_Street_Lighting_2010_1}, wireless sensor networks \citep{WSN_Street_Lighting_2010_1, WSN_Street_Lighting_2010_2} and both combined \citep{WSN_Street_lighting_2006, WSN_Street_lighting_2007}.

@@

% Recently, IETF has been standardized 6LoWPAN protocol (RFC4944) as an adaptation layer to enable IEEE 802.15.4-based low power and lossy networks (LLNs)

Recently, the 6LoWPAN protocol has been standardized by IETF as an adaptation layer to enable IEEE 802.15.4-based low power and lossy networks (LLNs), such as wireless sensor networks, to reuse existing IPv6 protocol for networking layer \citep{RFC4944}. The IP protocol is an open, mature and field proven technology, with free and well documented standards (RFCs), and thus it is very reasonable to reuse it for wireless sensor networking instead of creating new proprietary protocols. However, 6LoWPAN protocol does not define the routing protocol, which shall be defined apart. Thus, IETF has standardized RPL as the routing protocol for IP-based LLNs \citep{RFC6550}.

However, very recent empirical studies showed that RPL has some drawbacks, as higher amount of control messages than data messages, mainly for constructing routes from the sink to the nodes (downward routes in RPL vocabulary) \citep{Heurtefeux_RPL_2013,P2P_analysis_2010,P2P-RPL_2011}. By the other hand, in a street light system, nodes are static and a routing protocol could benefit of this condition to find routes with less control overhead than RPL. The routing algorithms that use position to find routes in a communication network are known as geographic routing algorithms and a large body of literature has been devoted to the research on those algorithms in last few decades \citep{Karp_GPSR_2000, Position_Based_Routing_Ad_Hoc_2001, On_Optimal_Geographic_Routing_2007, Denardin2011}.

Our main contribution in this paper is to propose a modification into RPL protocol, in order to better support P2P communication. The proposed technique intend to integrate a geographic routing algorithm in the RPL protocol to reduce the amount of control message required for P2P communication. We further propose a smart street lighting system based on IPv6-enabled wireless sensor network, which makes use of the new routing algorithm for message exchanging. This paper is organized as follows: Section \ref{sec6LoWPAN} presents an overview of the 6LoWPAN protocol, describing succinctly its operating modes, routing strategies and main drawbacks. In section \ref{secGeographicRouting} we address the geographic routing technique, highlighting its pros and cons. In section \ref{secGeoRank}, the \textit{GeoRank} algorithm is presented, depicting its working principle. We evaluate, by simulation, the performance of the proposed algorithm in Section \ref{secSimulation}. Finally, we conclude this paper in Section \ref{secConclusion}.

@@

6LoWPAN is a protocol definition, standardized in RFC 4944 \citep{RFC4944} and RFC 6282 \citep{RFC6282}, that works as an adaptation layer between the IPv6 network layer and the IEEE 802.15.4 MAC (Media Access Control) layer, as depicted in figure \ref{stack_6lowpan}. It is required because IEEE 802.15.4 standard, which defines MAC and PHY layers for low power and low data rate wireless networks, is able to carry packets with maximum size of 127 bytes, while IPv6 packets requires a MTU (minimum transmission unit) of 1280 bytes. Thus, the overall goal of 6LoWPAN is to allow IPv6 packets to be carried on top of IEEE 802.15.4-based wireless sensor networks, by defining fragmentation and compression mechanisms of the IPv6 packets and thus enabling IPv6-based wireless sensor networks \citep{Mulligan_6LoWPAN_2007}.

@@

\subsection{Routing in 6LoWPAN}
In order to address the lack of a routing protocol for 6LoWPAN networks, the RPL protocol has been proposed in RFC 6550 by IETF. Accordingly, RPL provides support for point-to-point (P2P), point-to-multipoint and multipoint-to-point traffic flows \citep{RFC6550}.

The RPL protocol is based on directed acyclic graph (DAG). DAGs are directed graphs formed by a collection of $n$ vertices labeled from the set \{1,2,...,n\}, such that the presence of the edge $(i,j)$ implies that $i < j$. Therefore, each directed edge connects one vertex to another, such that there is no way to start at some vertex $v$ and follow a sequence of edges that eventually loops back to $v$ again \citep{DAGS_1992}. All edges are contained in paths oriented toward and terminating at one or more root nodes. These root nodes are known as Low power and Lossy Network Border Routers (LBR), which are devices that connects the LLN to another routing domain such as a Local Area Network (LAN) or the internet, where a different routing protocol may be in operation. In addition to its functionality as a routing device, the LBR acts as a sink node to a WSN as well as a data collector or aggregator \citep{RFC7102}.

In RPL, a DAG grounded at a single LBR is called a Destination Oriented Directed Acyclic Graph (DODAG). For a specific DODAG each node is assigned to a rank such that rank increases as the node is far from the LBR. Note that the rank of the nodes along any path to the LBR must be monotonically decreasing, avoiding routing loops, as shown in the figure \ref{RPL_DIO}. Therefore, forwarding a packet to the LBR roughly consist in choosing the neighbor node with the lowest rank.

\begin{figure}[h!tb]
\centering
\includegraphics[width=1\textwidth]{figures/RPL_DIO}
\caption{Example of a Destination Oriented Directed Acyclic Graph(DODAG).}
\label{RPL_DIO}
\end{figure}

Nodes running RPL exchange control messages to setup and maintain the DODAG, which in the 6LoWPAN implementation are ICMPv6 messages. The 1 byte code in the ICMPv6 header is used to differentiate between different sub-type of control messages \citep{RFC4443}. The sub-type used to build the DODAG is called DAG Information Object (DIO) \citep{RFC6550}. All the nodes in the network regularly broadcast a DIO to its one-hop neighbors, indicating their activity and rank. The instant at which a node issues a DIO is governed by a Trickle timer \citep{trickle}, which is reset only when an inconsistency arises, i. e., very few DIOs are exchanged when the topology is stable. RPL provide optimized paths to one or more LBRs with very little information stored in the routing table. The information stored is limited to its one-hop neighbors and their parents in the DODAGs. A parent of a node within a DODAG is one of the immediate successors of the node on a path towards the DODAG root.

Despite of its advantages, the default RPL using DODAGs does not provide routes from the LBRs to other nodes in the network, which in RPL are known as down routes. Down refers to the direction from DODAG roots towards leaf nodes, following the reverse direction of DODAG edges. The availability of such routes is desirable in several scenarios, mainly in those with actuators nodes and where selective queries to sensors nodes are required. Thus, a RPL router that requires a path from a LBR to itself should send a DAO (Destination Advertisement Object) message to the LBR through the DODAG edges. The nodes that receive a DAO message forward the message as soon as the timer DelayDAO expires, being this time determined by the implementation, according to \citep{RFC6550}. Therefore, routes for any destination inside the network, other than the LBRs, are provided by these destinations through DAO messages.

@@

\begin{figure}[h!tb]
\centering
\includegraphics[width=1\textwidth]{figures/RPL_DAO}
%\includegraphics[width=0.45\textwidth]{figures/RPL_DAO}
\caption{P2P communication between routers $B$ and $C$ through a DODAG root using RPL.}
\label{RPL_DAO}
\end{figure}

@@

Even, there are problems related to the DAO operating modes. In non-storing mode, RPL requires the use of IPv6 source routing mechanism. Therefore, a Source Routing Header (SRH) for use strictly between RPL routers in the same RPL routing domain was proposed in RFC 6554. However, RPL source routing implements a policy where each IPv6 hop between the source and destination of the source route is specified within the SRH. Such approach leads to a large overhead and increases the consumption of the already scarce bandwidth. Further, as the IEEE 802.15.4 standard has a limited maximum transmission unit (MTU), that may also imply in increased fragmentation. Therefore, successful delivery of an IP packet depends on the successful delivery of all possible fragments, which are not desirable once the LLNs are generally characterized by high packet loss rates.

In addition to the fragmentation issue, the maximum length of the routing header is limited to 255 octets \citep{RFC4944}, which restricts the maximum path length to 127 hops if an address compression mechanism is applied \citep{herberg}. This limitation may exclude scenarios with long chain-like topologies, such as street lighting systems and traffic lights along a street.

In storing mode, each router has to be able to store all the routes downwards its position in the formed DODAGs. This implies that, for routers near to the root node, the required storage is only bounded by the number of paths to all other routers in the network \citep{RPL_observations}. As RPL targets networks consisting of thousands of memory-constrained devices, the required storing capacity on these devices may not be compatible with the basic precepts of the WSANs.

@@

Geographic routing mainly relies on a simple greedy geographic forwarding strategy, where each router node must select a locally optimal neighbor with a positive progress towards the data packet destination \citep{survey_2007}. Such approach is largely known by its low overhead and scalability, which can easily be observed due to its low control message system, i.e., a router in a geographic routing strategy only needs to know its direct neighborhood.

@@

\section{GeoRank}
\label{secGeoRank}

Mainly due to its greedy step, GOAFR algorithm is able to perform quite well in networks with high link density, discovering paths often optimal or near-optimal paths, in contrast to RPL, which is constrained to route only through the DODAGs and thus, being unable to find the optimal paths. By the other hand, in networks with low link density, mainly around the critical density \citep{GOAFR_2008}, GOAFR has a performance loss and may usually find paths longer than RPL. Therefore, a hybrid approach that leverages the best of both algorithms according to the network link density is proposed. This approach is mainly suitable for large scale and heterogeneous networks, such as a network of wireless street light nodes, where the link density varies from low to high depending of the more or less presence of obstructions and buildings in different places of the network.  The operation of \textit{GeoRank} is illustrated in figure \ref{state_machine} and the algorithm is outlined below in Algoritm \ref{alg}.

\small
\begin{algorithm}
\begin{algorithmic}
\caption{GeoRank}
\label{alg}
\small
\STATE \textbf{Step 1} - Store in the packet the DODAG root with the smallest angle to the source-destination
line. 
\STATE \textbf{Step 2} - Start with greedy step of GOAFR until: (case I) finding a void node (i.e. a node where GOAFR would switch to face  routing (OAFR)) or (case II) delivering the packet. In case I, store the current position in the packet and go to step 3.
\STATE \textbf{Step 3} - If the destination node is closer to the DODAG root (stored in the packet) than the current node or the DODAG root is closer to the destination node than the current node, forward to the next hop in the DODAG towards the root (i.e. upward route). If neither condition is satisfied, go to step 4. Whenever the current node position is closer to destination than the position stored in the packet in step 2, go to step 1.
\STATE \textbf{Step 4} -  Execute face routing mode of GOAFR until delivering the packet or returning to step 1.
\small
\end{algorithmic}
\end{algorithm}
\normalsize
\begin{figure}
\centering
\begin{tikzpicture}
[scale=1, every path/.style={>=latex}, no/.style={draw,circle,align=center,text width = 1.3cm}]
  \node[no] (a) at (0,0)  {\small GOAFR greedy routing};
  \node[no] (b) at (3,4)  {\small DODAG upward routing};
  \node[no] (c) at (6,0) {\small GOAFR face routing };

@@

For every scenario, there were generated $1000$ random source-destination pairs, giving a coverage of up to $10$\% of the possible source-destination combinations. Also, every simulation was performed for $10$ different (random) positions of the DODAG root in every scenario, which gives a good coverage of the $5$ possible root placements (i.e. center and top/bottom-right/left corners). Then, for the each source-destination pair, the route was determined by each algorithm under test (RPL, GOAFR, \textit{GeoRank} and Dijkstra' algorithm for shortest path) and length of each route was measured in number of hops. The average length of the routes was computed and the results are shown in figure \ref{sim_results}.
%For RPL rank computation, it was used the Objective Function 0 as defined in RFC6552.
\begin{figure}
\centering
\includegraphics[width=1\textwidth]{figures/sample_scenarios.eps}
%\includegraphics[width=0.45\textwidth]{figures/sample_scenarios.eps}
\caption{Sample of simulated scenarios with nodes populated.}
\label{sim_scenarios}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{figures/final_result.eps}
%\includegraphics[width=0.45\textwidth]{figures/final_result.eps}
\caption{Simulation results of the average route length for the routing algorithms (RPL, GOAFR, \textit{GeoRank}, and Dijkstra' algorithm for shortest path).}
\label{sim_results}
\end{figure}

@@

\bibliographystyle{bst_models/model2-names}
\bibliography{refs}
